---
title: "Week 4 Lab: Geoms and Aesthetic Mappings"
author: "Daniel Lip, Pierre Chan, Howard Yeo, Joel Koh, Jolie Yeo, Danish Dol, Sarah Loo"
format: html
editor: visual
---

# 1 Visualizing the Distribution of Travelers on the Titanic

## 1.1 Bar Chart of Travelers by Class

```{r}
#| label: Task 1.1

library(tidyverse)
library(ggplot2)

titanic_cat <- function(...) {
  ggplot2::labs(
    x = "",
    y = "",
    title = "Titanic Travelers by Class",
    caption = 'Source: Encylopedia Titanica (2023)',
    ...
  )
}

titanic_data <- read_csv("titanic.csv")
ggplot(titanic_data, aes(y=class)) +
  geom_bar(width=0.5,fill=NA, color = "black") +
  geom_text(stat = "count", aes(label = ..count..), hjust = -0.2) +
  titanic_cat() +
  theme(
    axis.text.x = element_blank(),       
    axis.ticks.x = element_blank(),      
    plot.title = element_text(hjust = 0.5)
  ) +
  xlim(0, 950) 

```

## 1.2 Dodged Bar Chart of Travelers by Class and Survival

```{r}
#| label: Task 1.2
ggplot(titanic_data, aes(y = class, fill = survived)) +
  geom_bar(position = "dodge") +
  geom_text(
    aes(label = after_stat(count), color = survived),
    stat = "count",
    position = position_dodge(width = 0.9),
    hjust = -0.2,
    size = 3,
    fontface = "bold"
  ) +
  titanic_cat(fill = "Survived") +
  labs(title = "Titanic Travelers by Class and Survival") +  # Overriding title here
  coord_cartesian(clip = "off") +
  scale_fill_manual(values = c("FALSE" = "salmon", "TRUE" = "turquoise")) +
  scale_color_manual(values = c("FALSE" = "salmon", "TRUE" = "turquoise")) +
  guides(color = "none")
```

## 1.3 Mosaic Plot

```{r}
library(ggmosaic)
ggplot(titanic_data) +
  geom_mosaic(aes(x= product(survived, class), fill=survived))+
  scale_fill_manual(values = c("#F46D43", "#FEE090"))+
  theme_mosaic()+
  labs(
    y="Survived",
    x="Class",
    title="Titanic Travelers by Class and Survival"
  )
```

## 1.4 Pros and Cons of Mosaic Plots

An advantage that mosaic has over the dodged bar chart would be that it shows a clearer visual representation of the proportion and total. Making it easier to assess the class size and survival proportion at the same time as the width of each class in the plot reflects the total number of people in that class.

A disadvantage would be that as mosaic plot are generally more complex, it is harder to do a comparison unlike the dodged bar chart which provide a clear side by side comparison of survival count.

## 1.5 Alternative Visualization

```{r}
#| label: heatmap-class-survival
#| message: false
titanic_data %>%
  count(class, survived) %>%
  mutate(
    class = factor(class, levels = c("1st", "2nd", "3rd", "Crew")),
    survived = factor(survived, levels = c("FALSE", "TRUE"))
  ) %>%
  ggplot(aes(x = class, y = survived, fill = n)) +
  geom_tile(color = "white") +
  geom_text(aes(label = n), size = 4) +
  scale_fill_gradient(low = "#FEE0D2", high = "#DE2D26") +
  labs(
    title = "Heatmap of Class and Survival Count",
    x = "Class",
    y = "Survived",
    fill = "Count",
    caption = "Source: Encyclopedia Titanica (2023)"
  ) +
  theme_minimal() +
  theme(
    plot.caption = element_text(hjust = 0.5),
    plot.title = element_text(hjust = 0.5),
    panel.background = element_rect(fill = "grey90", color = NA)
  )
```

### Discuss the advantages and limitations of your chosen visualization compared to bar and mosaic plots.

For Task 1.5, I chose a **heatmap** to visualize the relationship between class and survival. This method encodes the count of individuals using color intensity and overlays the exact values for clarity. The x-axis represents class, while the y-axis captures survival status (TRUE/FALSE).

**Advantages:**

-   Colour encoding provides for quick visual density cues, making it easy to spot which groups, in this case, 3rd class non-survivors, had the highest counts due to stronger colour intensity.

-   Direct number labeling avoids ambiguity and supports precise interpretation.

-   The heatmap provides a compact visual summary and shows all combinations in a single, space-efficient layout.

**Limitations:**

-   Compared to a proportional bar chart, the heatmap does not convey survival rates within each class, only absolute counts, to show inequality across classes.

-   The mosaic plot conveys proportions using area, which is conceptually powerful, but can be harder to interpret accurately, especially for unfamiliar users.

## 2 Find a Data Set to Demonstrate Semantically Correct Aesthetic Mappings

## 2.3 Task

a.  Provide the data URL. Data Source: https://vincentarelbundock.github.io/Rdatasets/doc/datasets/mtcars.html

b.  Briefly summarize the scope, origin, and key variables of your data set in your own words. The **mtcars** dataset was extracted from the 1974 US magazine Motor Trend. It provides specifications and performance metrics for 32 different car models from the early 1970s. Key variables include:

-   mpg: Miles per gallon (fuel efficiency)
-   wt: Weight of the car (in 1000 lbs)
-   hp: Gross horsepower
-   gear: Number of forward gears (categorical)

This dataset allows for exploration of how physical characteristics e.g. weight and horsepower relate to fuel efficiency.

c.  Use ggplot2 (and any necessary add-on package) to produce a single plot incorporating at least two out of three aesthetic mapping categories listed in Section 2.1. Choose any plot type that best illustrates your variables.

```{r}
#| label: Task 2.3

motor_trends <- read_csv("mtcars.csv")

ggplot(motor_trends, aes(x = wt, y = mpg, color = factor(gear), size = hp)) +
  geom_point(alpha = 0.8) +
  labs(
    title = "Fuel Efficiency vs Weight of Car",
    x = "Weight (1000 lbs)",
    y = "Miles per Gallon",
    color = "Number of Gears",
    size = "Horsepower",
    caption = "Source: US EPA dataset via Rdatasets"
  ) +
  theme_minimal()

```

d.  For each aesthetic mapping, explain why it is appropriate, referencing the variable nature and perceptual best practices. For example: “We mapped the population variable to the size aesthetic because the variable is extensive: when the sample size changes, the expected value of population also changes by the same factor.”

-   **Color**: The `gear` variable is categorical, representing the number of forward gears in the car. Using color to differentiate these categories allows for easy visual distinction between groups.
-   **Size**: The `hp` (horsepower) variable is quantitative and extensive, meaning that as horsepower increases, the expected performance of the car also increases. Mapping horsepower to size allows viewers to quickly gauge the power of each car model relative to its weight and fuel efficiency.

e.  State the main insight or conclusion the reader should draw from your visualization. Cars with fewer gears and higher horsepower generally have lower fuel efficiency and higher weight. In contrast, lighter cars with more gears tend to be more fuel-efficient. This suggests a trade-off between engine power and mileage performance.
