---
title: "Week 4 Lab: Geoms and Aesthetic Mappings"
author: "Daniel Lip, Pierre Chan, Howard Yeo, Joel Koh, Jolie Yeo, Danish Dol, Sarah Loo"
format: html
editor: visual
---

# 1 Visualizing the Distribution of Travelers on the Titanic

## 1.1 Bar Chart of Travelers by Class

```{r}
#| label: Task 1.1

library(tidyverse)
library(ggplot2)

titanic_cat <- function(...) {
  ggplot2::labs(
    x = "",
    y = "",
    title = "Titanic Travelers by Class",
    caption = 'Source: Encylopedia Titanica (2023)',
    ...
  )
}

titanic_data <- read_csv("titanic.csv")
ggplot(titanic_data, aes(y=class)) +
  geom_bar(width=0.5,fill=NA, color = "black") +
  geom_text(stat = "count", aes(label = ..count..), hjust = -0.2) +
  titanic_cat() +
  theme(
    axis.text.x = element_blank(),       # Hide x-axis text
    axis.ticks.x = element_blank(),      # Hide x-axis ticks
    plot.title = element_text(hjust = 0.5)  # Center the title
  ) +
  xlim(0, 950) 

```

======= \## 1.2 Dodged Bar Chart of Travelers by Class and Survival

```{r}
#| label: Task 1.2
ggplot(titanic_data, aes(y = class, fill = survived)) +
  geom_bar(position = "dodge") +
  geom_text(
    aes(label = after_stat(count), color = survived),
    stat = "count",
    position = position_dodge(width = 0.9),
    hjust = -0.2,
    size = 3,
    fontface = "bold"
  ) +
  titanic_cat(fill = "Survived") +
  labs(title = "Titanic Travelers by Class and Survival") +  # Overriding title here
  coord_cartesian(clip = "off") +
  scale_fill_manual(values = c("FALSE" = "salmon", "TRUE" = "turquoise")) +
  scale_color_manual(values = c("FALSE" = "salmon", "TRUE" = "turquoise")) +
  guides(color = "none")
```

## 1.3 Mosaic Plot

```{r}
library(ggmosaic)
ggplot(titanic_data) +
  geom_mosaic(aes(x= product(survived, class), fill=survived))+
  scale_fill_manual(values = c("#F46D43", "#FEE090"))+
  theme_mosaic()+
  labs(
    y="Survived",
    x="Class",
    title="Titanic Travelers by Class and Survival"
  )
```

## 1.4 **Pros and Cons of Mosaic Plots**

An advantage that mosaic has over the dodged bar chart would be that it shows a clearer visual representation of the proportion and total. Making it easier to assess the class size and survival proportion at the same time as the width of each class in the plot reflects the total number of people in that class.

A disadvantage would be that as mosaic plot are generally more complex, it is harder to do a comparison unlike the dodged bar chart which provide a clear side by side comparison of survival count.
